---
title: "Appendices"
author: "Andrew Wolfe and Morgan Huhndorff"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r packages, include = FALSE}
# Load any packages you may need here
library(car)
library(emmeans)
```


```{r readData, include = FALSE}
#Read in the data 
cancer <- read.csv("cancer_mortality_rates.csv")
```

***

#### Figure 1: Numerical Summaries of Data

```{r figure1, echo = FALSE}
summary(cancer[c(1,3,14,17)])

lm_cancer <- lm(cmRate ~ medianIncome + pctPrivateHC + pctWhite, data = cancer)
summary(lm_cancer)
```

***

#### Figure 2: Scatterplot Matrix of Data

```{r figure2, echo=FALSE}
plot(~ cmRate + medianIncome + pctPrivateHC + pctWhite,data = cancer)
```

***

#### Figure 3: Histogram of Variables

```{r figure3, echo=FALSE}
#response
hist(cancer$cmRate,
     main = "Histogram of Cancer Mortality Rates",
     xlab = "Mortalities (per 100,000 people)")

#predictors
hist(cancer$medianIncome,
     main = "Histogram of County Median Income",
     xlab = "Median Income ($)")
hist(cancer$pctPrivateHC,
     main = "Histogram of Residents with Private Healthcare",
     xlab = "Percentage")
hist(cancer$pctWhite,
     main = "Histogram of Percentage of White Residents",
     xlab = "Percentage")
hist(cancer$pctEmployed,
     main = "Histogram of Percentage of Employed Residents",
     xlab = "Percentage")
```

***

#### Figure 4: Transformation of Percentage White Variable

```{r figure4, echo = FALSE}
cancer$pctNonWhite <- (100 - cancer$pctWhite) + 0.1
hist(log(cancer$pctNonWhite),
     xlab = "log(pctNonWhite)",
     main = "Histogram of log(pctNonWhite)")
```

***

#### Figure 5: Quadratic Term

```{r figure5, echo = FALSE}
lm_quadratic <- lm(cmRate ~ medianIncome + I(medianIncome^2), data = cancer)
summary(lm_quadratic)

```
*** 

### Figure 6: Scatterplot

```{r figure6, echo = FALSE}
plot(cmRate ~ pctEmployed, data = cancer,
     xlab = "Residents Employed (%)",
     ylab = "Cancer Mortalities (per 100k)",
     main = "Scatterplot of Cancer Mortalities vs Residents Employed"
     )
```


***

#### Figure 7: Categorical Variable Region

```{r figure7, echo = FALSE}
boxplot(cancer$cmRate ~ cancer$region,
        main = "Boxplots of Cancer Mortalities by Region",
        ylab = "Cancer Mortalities (per 100,000 people)",
        xlab = "Region")
```

```{r, echo = FALSE}
#Test for a difference in the estimated means of each Region
lm_r <- lm(cmRate ~ region, data = cancer)
pairs(emmeans(lm_r, specs = "region"), adjust = "none")
```

***

#### Figure 8: Interaction Term

```{r figure8, echo = FALSE}
#Change region to a factor
cancer$region <- as.factor(cancer$region)

lm_region <- lm(cmRate ~ pctPrivateHC * region, data = cancer)
summary(lm_region)
```
***

#### Figure 9: Scatterplot by group

```{r figure9, echo = FALSE}
plot(cmRate ~ pctPrivateHC,
     data = cancer,
     col = c("red","blue","yellowgreen","gray40","purple4")[region],
     pch = c(1,2,3,4,5)[region],
     xlab = "Residents with Private Health Coverage (%)",
     ylab = "Cancer Mortalities (per 100,000 people)",
     main = "Scatterplot of Cancer Mortalities vs Private Health Care \n by Region")

abline(a = 262.36621, b = -1.24555, col = "red")
abline(a = 262.36621 + 1.95485 , b = -1.24555 - 0.03282, col = "blue")
abline(a = 262.36621 - 20.92079, b = -1.24555 + 0.43423, col = "yellowgreen")
abline(a = 262.36621 - 95.70596, b = -1.24555 + 1.33823, col = "gray40")
abline(a = 262.36621 - 55.54333, b = -1.24555 + 0.49217, col = "purple4")

legend(x= 78, y = 300,
       legend = levels(cancer$region),
       col = c("red","blue","yellowgreen","gray40","purple4"),
       pch = c(1,2,3,4,5))
```
***

#### Figure 10: Model Comparison
```{r figure10, echo = FALSE}
lm_interaction <- lm(cmRate ~ medianIncome + I(medianIncome^2) + log(pctNonWhite) + pctEmployed + pctPrivateHC * region, data = cancer)
lm_noInteraction <- lm(cmRate ~ medianIncome + I(medianIncome^2) + log(pctNonWhite) + pctEmployed + pctPrivateHC, data = cancer)

summary(lm_interaction)
summary(lm_noInteraction)
```
***
#### Figure 11: Assumptions for Models

```{r figure11, echo = FALSE}
qqnorm(lm_interaction$residuals, main = "QQ Plot of the Residuals")
qqline(lm_interaction$residuals)

plot(lm_interaction$fitted.values, lm_interaction$residuals,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted Values Plot")

abline(h = 0, lty = 2, col = "red")
```

***

#### Figure 12: Summary Ouput of Model 1

```{r figure12, echo = FALSE}
lm_cancer <- lm(cmRate ~ medianIncome + I(medianIncome^2) + log(pctNonWhite) + pctPrivateHC + pctEmployed, data = cancer)
summary(lm_cancer)
```

***

#### Figure 13: VIFs for Model 1

```{r figure 13, echo = FALSE}
vif(lm_cancer)
```

***







